---
title: "Walkthrough1"
author: "Martijn Sturm"
date: "9 september 2019"
output: 
  html_document:
    df_print: paged
    fig_height: 6
    fig_width: 3
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidytext)
library(tidyverse)
library(xlsx)

# Import abbreviation list
import_abbs <- function(filename) {

  df <- read_excel(paste0("data/", filename)) %>%
    select(1:4)
  colnames(df) <- c("abbreviation", "expansion", "explanation", "extra")
  # Rows without a present expansion will not be used for replacement
  df <- df[!is.na(df$expansion), ]
  cat("\nImport of abbreviation list is completed")
  return(df)
}

### Creates a new column in the abbreviation dataframe
### This column coppies the abbreviations but strips them from a trailing '.'

# Sub function to actually remove the '.' from the string.
rm_trailing_dot <- function(x){
  string <- x
  # Regex: matches entire string which ends with a '.'
  pattern <- pattern <- "^.*\\.$"
  # Replaces matches with same string but last character stripped
  replacement <- substr(string, 1, nchar(string)-1)
  result <- sub(pattern = pattern, replacement = replacement, x = string)
  return(result)
}

# Function to replace abbreviation with trailing dots in string
# Necessary since 'unnest_token()' will strip trailing dots,
# which makes matching of tokens to abbreviations impossible otherwise.
replace_trailing_dot <- function(abb_df) {
  # Initialize counter to keep track of number of changes
  counter <- 0
  new_df <- abb_df
  new_df$abb_cor <- NA
  cat("\nReplacement of trailing '.' in abbreviations dataframa has been initiated: \n\n")
  for (i in seq_len(nrow(abb_df))) {
    abb <- abb_df$abbreviation[i]
    result <- rm_trailing_dot(abb)
    if (result != abb) {
      print(paste(abb, "has been replaced to", result))
      counter <- counter + 1
    }
    new_df$abb_cor[i] <- result
  }
  cat("\n")
  cat(paste("A total number of", counter, "replacements for trailing '.' has been made.\n\n"))
  return(new_df)
}

# # Redundant:
# # Function to replace abbreviations in token filled dataframe
# replace_abb <- function(token_df, abb_df) {
#   new_df <- token_df
#   for (i in seq_len(nrow(token_df))) {
#     word = token_df$word[i]
#     replacement <- expand_abb(word, abb_df)
#     new_df$word[i] <- replacement
#   }
#   return(new_df)
# }

# Function to match expansion with abbreviation
expand_abb <- function(word, abb_df) {
  # Check every record in the abbreviation list
  for (i in seq_len(nrow(abb_df))) {
    # If the abbreviation in abbreviation list matches the token from patientrecord:
    if (word == abb_df$abb_cor[i]) {
      # return the the abbreviation's expansion
      result = abb_df$expansion[i]
      return(result)
    }
  }
  # If no match in the abbreviation list, return the original word
  result = word
  return(result)
}

# Function to keep token filled dataframe, and add abbreviation expansion information
add_expansion <- function(token_df, abb_df) {

  cat("\nDataframe with tokens (words) is being scanned for known abbreviations...\nThis can take some time\n")

  # Copy dataframe to new dataframe which will be function's output
  new_df <- token_df %>% mutate(expansion = NA, abb_cor = NA, converted = NA)
  # Loop through every token generated from patient's token_df
  for (i in seq_len(nrow(token_df))) {
    word = token_df$word[i]

    # Get token's expansion, if token is no abbreviation, token itself will be returned
    expansion <- expand_abb(word, abb_df)

    if (expansion != word) {
      new_df$expansion[i] <- expansion
      new_df$abb_cor[i] <- word
      new_df$converted[i] <- TRUE
    }
    else {
      new_df$expansion[i] <- word
      new_df$abb_cor[i] <- NA
      new_df$converted[i] <- FALSE
    }
  }
  # Create an overview of every token with its abbreviation list information if present
  new_df <- left_join(new_df, abb_df)
  # Count the number of conversions: abbreviation -> expansion
  nr_conv <- new_df %>% filter(converted == TRUE) %>% nrow()
  cat("\nScan for abbreviations is completed.\n")
  print(paste("Total number of tokens (words) that have been converted from abbreviated to expanded form is: ", nr_conv))
  return(new_df)
}

# Function to expand abbreviations in the complete unedited patient records
expand_text <- function(pat_records, token_df) {
  mismatch_counter <- 0
  cat("\nReplacement of abbreviations to expansions in patient records is initiaded\n")
  # Check if the number of records between abbreviation dataframe and original records are the same
  if (all.equal(unique(pat_records$record), unique(token_df$record))) {
    cat("\nPatient records and generated abbreviation dataframe are in accordance with each other\n")
  } else {
    cat("\nProblem with number of records between patient records and abbreviation list\n")
    return()
  }

  # Only the abbreviations that were recognized in patient records are necessary to keep in and loop through
  # All tokens that were not recognized as abbreviations can be filtered out:
  token_df <- token_df %>% filter(converted == TRUE)

  # Go over the records one by one to change abbreviations into expansion
  for (i in seq_len(nrow(pat_records))) {

    # Select the abbreviations from the token dataframe for the record that is currently looped over
    record_nr <- pat_records$record[i]
    token_df_rec <- token_df %>% filter(record == record_nr)

    # Select the record from patients record that needs to be changed
    record <- pat_records$text[i]

    # Loop over every known abbreviation from the current patient record, and change it into its expansion
    for (j in seq_len(nrow(token_df_rec))) {

      # Creating a regex pattern to match the current abbreviation that is looked for
      # "(?i)" is case insensitivity flag. (\\W) are non-word character lookarounds (behind and ahead) (\\b) is word bounary
      match <- paste0("(?i)(?<=(\\W)|(\\b))", token_df_rec$word[j], "(?=(\\W)|(\\b))")

      # replacement is the corresponding expansion from the token dataframe
      replacement <- token_df_rec$expansion[j]
      #print(paste("replacement:", replacement))

      # Ignore case since abbreviation capital letter usage can be inconsistent
      change <- str_replace(pattern = match, replacement = replacement, string = record)
      #print(paste("change", change))

      # Report to user if the abbreviation was not recognized in the record during this instance of the loop
      found_match <- str_detect(string = record, pattern = match)
      #print(found_match)
      if (found_match == FALSE) {
        mismatch_counter <- mismatch_counter + 1
        print("This abbreviation was not recognized in the patient record: ")
        print(paste("abb:", token_df_rec$word[j], "to expansion:", replacement, "in record:", token_df_rec$record[j]))
      }

      # Keep change in record variable
      record <- change
    }
    # Put edited record back into dataframe
    pat_records$text[i] <- record
  }

  if (mismatch_counter > 0) {
    cat("\nSome abbreviations could not be replaced:\n")
    print(paste("Number of mismatches:", mismatch_counter))
  }
  else {
    cat("\nAll identified abbreviations were replaced by their abbreviations.\n")

  }
  cat("\nAbbreviation replacement has been completed.\n")
  return(pat_records)
}
```

## Abbreviation module
In this module we try to replace abbreviations in patient records  

1. Patient records and the abbreviation list are read from excel files into R dataframes
2. Each patient record will be split up into words. This is called tokenization.
3. Modification of abbreviations. Trailing dots need to be removed from the abbreviations
4. Scan patient records for abbreviations in abbreviation list
5. Expand the abbreviations in-text
6. Export results

### Import data and abbreviations
```{r}
# Import patient records
pat_records_orig <- read_excel("data/test_data2.xlsx") %>% na.omit()
# Display head
head(pat_records_orig)
```

The function to import abbreviations removes rows which have an empty cell in the 'expansion' column.
```{r}
# Import abbreviation list
abb_df <- import_abbs("abb1.xlsx")
head(abb_df)
```

### Tokenization of patient records
Tokenizer:  

* Lowers tokens (no capitals anymore)
* Removes trailing dots ('.') at end of words, assuming those are full stops for sentences

```{r}
# Transform text into a dataframe with tokens per word
token_df <- pat_records_orig %>% unnest_tokens(word, text) #to_lower = False, can be used to keep capitalization
token_df
```


### Remove trailing dots from abbreviations
Since the tokenizer gets rid of trailing dots on the tokens, abbreviations that end with a dot won't be recognized.
Therefore, the trailing dots need to be removed from the abbreviations in the abbreviation list.

```{r}
abb_df <- replace_trailing_dot(abb_df)
abb_df
```

### Scan for abbreviations in records (add_expansion())
This function loops through the tokenized patient records to check if the token matches one of the abbreviations.
It returns a dataframe based on the token_df above. But now information is added whether each token is recognized as abbreviation, and which expansion corresponds to it.  

```{r}
token_df <- add_expansion(token_df, abb_df)
token_df
```

This Dataframe will be used to replace the abbreviations in-text in the patient records

### Expand abbreviations in text (expand_text())
This function uses Regex to replace the abbreviations identified upstream, by their replacements

```{r}
pat_records_exp <- expand_text(pat_records = pat_records_orig, token_df = token_df)
pat_records_exp
```

### Export to Excel
Files can be exported to Excel to be made accessible for others.
Or to use it as input for another downstream task in Python.

```{r}
# Writes token dataframe to an excel file in the output folder
token_df %>% filter(converted == TRUE) %>% as.data.frame() %>%
  write.xlsx(file = "output/attempt1_abbs.xlsx", col.names = T, row.names = F)
```


```{r}
# Writes adjusted patient records to an excel file in the output folder
pat_records_exp %>% as.data.frame() %>%
  write.xlsx(file = "output/attempt1_pat_records.xlsx", col.names = T, row.names = F)
```

